\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-18 pldi/draft.tex
a couple of remarks

A node $f$ of the graph represents a function from a (possibly empty)
list of input streams to a list of output streams $f : [D\kstar] \to
[D\kstar]$, where $D$ represents the basic data type of a line of
characters. This representation captures the majority of shell
commands. We require that the function $f$ is monotone w.r.t.\ a
lifting of the prefix order for a sequence of inputs \kk{Formalize
  this lifting if essential for later development}.
%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
%% concatenation and will be omitted when obvious.
This captures the idea that a node cannot retract output that it has
already produced.

@@

\kk{Find a better name.} \todo{One-input-output} commands \todo{are
  special commands} that consume their inputs one at a time, with the
possible exception of static input files, and produce one output
file. These commands can be represented as functions of type $f :
D\kstar \times [D\kstar] \to D\kstar$, where the first argument
represents the concatenation of the inputs that are consumed one at a
time and the second argument represents the static inputs. The
quintessential example of \todo{one-input-output} commands is
\ttt{cat}, that consumes its inputs one at a time, producing their
concatenation as output. We focus on these commands because a large
subset of \sta and \pur commands fall in this category.

@@

As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
operate independently on individual elements of the stream
(characters, lines, or files) without maintaining any state. In this
work, we consider lines as the data quantum, so only commands that are
stateless regarding lines (or any finer granularity such as
characters) are considered stateless. Formally, a
\todo{one-input-output} command $f$ is stateless if it is ``linear''
with respect to concatenation, \ie satisfying the following equation:

\[
\forall x, x', s, f(x.x', s) = f(x, s).f(x', s)
\]

@@

\[
\forall x, x', s, f(x.x', s) = f'(m(x, s).m(x', s), s)
\]

\noindent
This means that we get the same output by applying $f$ to a
concatenation of two inputs $x, x'$, or by applying the fold
function $f'$ to the concatenation of the outputs produced by applying
map $m$ to each of these inputs.

\heading{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

>>>>

A node $f$ of the graph represents a function from a (possibly empty)
list of input streams to a list of output streams $f : [D\kstar] \to
[D\kstar]$, where $D$ represents the basic data type of a line of
characters. \km{Is the number of input/output streams variable? If $m$ is the number of input streams, is the number of output streams a function of $m$? If this the case, then it is not captured by this type. Maybe it would help to show a couple of example operators together with the types of their denotations. A graphical representation (as in Fig.~\ref{fig:parallelization-transformation} would also be nice.} This representation captures the majority of shell
commands. We require that the function $f$ is monotone w.r.t.\ a
lifting of the prefix order for a sequence of inputs \kk{Formalize
  this lifting if essential for later development}.
%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
%% concatenation and will be omitted when obvious.
This captures the idea that a node cannot retract output that it has
already produced.

@@

\kk{Find a better name.} \todo{One-input-output} commands \todo{are
  special commands} that consume their inputs one at a time, with the
possible exception of static input files, and produce one output
file. These commands can be represented as functions of type $f :
D\kstar \times [D\kstar] \to D\kstar$, where the first argument
represents the concatenation of the inputs that are consumed one at a
time and the second argument represents the static inputs. \km{Again, a graphical presentation of these operators would be useful. Maybe with a way to distinguish between between streaming inputs and static inputs? Btw, this distinction is not completely clear to me at this point. Clearly, not all programs that we can write are streaming (monotone).} The
quintessential example of \todo{one-input-output} commands is
\ttt{cat}, that consumes its inputs one at a time, producing their
concatenation as output. We focus on these commands because a large
subset of \sta and \pur commands fall in this category.

@@

As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
operate independently on individual elements of the stream
(characters, lines, or files) without maintaining any state. In this
work, we consider lines as the data quantum, so only commands that are
stateless regarding lines (or any finer granularity such as
characters) are considered stateless. Formally, a
\todo{one-input-output} command $f$ is stateless if it is ``linear''
with respect to concatenation \km{It commutes with the operation of concatenation. This is the definition of (semigroup) homomorphism. I am not sure that linearity is an appropriate term.}, \ie satisfying the following equation:

\[
\forall x, x', s, f(x \cdot x', s) = f(x, s) \cdot f(x', s)
\]

@@

\[
\forall x, x', s, f(x \cdot x', s) = f'(m(x, s) \cdot m(x', s), s)
\]

\noindent
This means that we get the same output by applying $f$ to a
concatenation of two inputs $x, x'$, or by applying the fold
function $f'$ to the concatenation of the outputs produced by applying
map $m$ to each of these inputs. \km{We had also discussed here the case of operators that can be described as associative and commutative functions on the data (e.g., sum). In this case the parallelization has the flavor of data parallelism: one can split the input stream in an arbitrary way (e.g., randomly or using some other load-balancing scheme) and merge the outputs. For the more broad class of associative operations (e.g., sort), the formulation of this paragraph makes perfect sense. I just want to mention that this feels a lot more like \textbf{batching} than map-reduce-like data parallelization.}

\heading{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s) \cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the behavior of the graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s) \cdot m(x_2, s) \cdots m(x_n, s), s)$. A visual representation of this transformation can be seen in \Cref{fig:parallelization-transformation}.
